World Heritage Sites on Wikipedia: Controversies in a Context of Constrained Agency for Cultural Heritage Activism

### 2020-06-01 analysis/paper/paper.Rmd
submission to BD&S

An additional consideration for understanding participation in online peer production communities are the technical schemas of the software that Wikipedia runs on, MediaWiki. This is a complex toolkit that enables participation in Wikipedia in highly structured ways. On one hand, these structured behaviours produce structured datasets that are well suited to data science methods for efficient computational analysis of large numbers Wikipedia articles. On the other hand, they constrain and limit the agency of the user, canalizing their behaviour into a small number of possible actions and acceptable modes of discourse and engagement with other users [@iba2010analyzing]. While Wikipedia has elements that are ubiquitous on the Internet, such as links on articles that take the user to other articles or other pages on the Internet, it also has several less common elements that contribute to its unique technicity, resulting in very specific types of relationships between human users and the technical elements of the Wikipedia project [@niederer2010wisdom; @weltevrede2016platform]. For example, every edit to an article is tracked in a publicly accessible version control system attached to that article. This exposes the article creation process in highly granular detail; for any given article, we can see how many editors contributed, the size of their edits and their distribution over time, among other things [@priedhorsky2007creating]. Wikipedia has a special category of edit called the 'revert' which allows a user to restore an article to an earlier state to remove recent vandalism (such as the addition of irrelevant or offensive material). This special revert action, combined with a 'talk' page attached to each encyclopedia article for threaded discussion among editors, allows us to detect and study the dynamics of conflicts arising from the creation and editing of articles [@suh2007us]. For example, a high proportion of 'revert' type edits, and a high word count on the talk page can indicate the controversiality of an article [@yasseri2012dynamics].

@@

Of the `r nrow(wh_unesco)` cultural sites on the World Heritage List at the time of writing, we found Wikipedia articles for `r nrow(some_page_variables)`. To characterize the overall variability of article according to the basic characteristics described above, we applied a Uniform Manifold Approximation and Projection (UMAP) to reduce the dimensionality of our data [@mcinnes2018umap]. After a two-dimensional projection was obtained, we used Hierarchical Density-Based Spatial Clustering of Applications with Noise (HDBSCAN) to identify clusters [@campello2013density], and Random Forests to identify variable loadings [@breiman2001random] so we could determine which of the basic characteristics were most important in driving variability between the articles. This revealed `r n_clusters` clusters, with the most important variables for determining cluster membership being the number of items in the reference list at the end of the article, the number of words in the article, and the number of Wikilinks out from the target article to other Wikipedia articles. That said, `r n_no_cluster` (`r round(prop_no_cluster) * 100`%) articles were not assigned to a cluster by the HDBSCAN algorithm, indicating that the basic characteristics of CS-WHL articles do not partition the articles in highly structured and discrete groups.

@@

The scatterplot on the lower right of Figure \@ref(fig:wh-wiki-content-and-consumption) shows the values of inward Wikilinks, article views and article word count for all CS-WHL articles. The labeled points in the upper right quadrant are the articles that receive most of the attention. Of the CS-WHL notable for conflict discussed above, only Timbuktu is visible among these highly popular articles. It is also the only site on the List of World Heritage in Danger that is among these highly popular articles. We reviewed the talk page for the Timbuktu article to determine if the attention received by the article might relate to the issues leading to armed conflict at the site. Of the `r timbuktu_talk_page_word_count` word on the talk page, only 168 are on the topic of conflict and destruction, with the editors discussing how to describe the scale of the damage to the temples. The majority of talk page content is about filling in missing detail, suggestions for or notifications of minor corrections. This is also the case for the talk page for Auschwitz, another very popular CS-WHL article. Although Wikipedia articles on this site in other languages have different points of emphasis [@wolniewicz2016framing], the discussion among editors is generally technical and precise, and provocative comments usually ignored rather than directly disputed. This can be contrasted with the talk page for the Holocaust, where @pfanzelter2015crossroads found abundant evidence of conflict among editors.

@@

For the special 'revert' edit type, we see that the proportion of all edits per  CS-WHL articles is similar to other articles, but has a left-skewed distribution indicating a higher number of articles that have very few revert edits (Figure \@ref(fig:wh-wiki-vandal-revert)). We also identified edits with the string 'vandal' in the edit summary as a similar type of edit to the revert edit, e.g. "Edits by 72.49.241.71 identified as vandalism". CS-WHL articles generally have fewer edits about vandalism than our random sample. The shape of the distribution of edits about vandalism has a smaller second mode to the left of the peak, indicating that a large number of CS-WHL articles have very few edits about vandalism (Figure \@ref(fig:wh-wiki-vandal-revert)). Among the CS-WHL articles that have high proportions of reverts and edits about vandalism are highly iconic sites in the Western canon of culture history, e.g. the Sydney Opera House, the Tower of London, and the Statue of Liberty. In reviewing a sample of several hundred reverted edits for each of these, we found that nearly all of them are undoing the addition of short strings of text (e.g. profanities, spam, nonsense). Much of this vandalism is playful, in the spirit of "'I am', a statement that one is present and alive", as @baker2003red described historical graffiti on the Reichstag in Germany by Russian soldiers in WW2. Once again, of the CS-WHL sites with a history of conflict or on the in-danger list, only Timbuktu appears here as having high proportions of revert and vandalism-reversing edits.

@@

Perhaps the most striking spatial pattern in our results is how much of the edits to CS-WHL articles in any part of the world come from the United States. The anonymous editors of articles about CS-WHL are generally not located in the same country as the site they are editing. This suggests a digital colonialism of world heritage information on Wikipedia, where heritage is interpreted and communicated by people who are not part of the descendant community traditionally associated with the site. This is an important observation because it is these editors that control the production of knowledge about CS-WHL articles, and the ability to control information about heritage has several implications. First is that the choices of these editors define community identity, that is, the community of Wikipedia editors, and also the broader community of Internet users that read Wikipedia, by indicating which sites are important and visible, and which are not by contributing to a digital 'authorized heritage discourse' [@smith2012constrained] that prioritizes their own self-interests. Second is that traditional communities associated with CS-WHL have little control over how their own heritage is represented on Wikipedia. We see this spatial pattern on Wikipedia as part of a trend of globalization of heritage, where places that had traditionally contributed to local or regional identities are turned into sanitized playgrounds for rich tourists [@bernbeck2004political]. Wikipedia may be seen as virtual playground for Americans to construct a universal heritage that reflects their neo-liberal cultural values, and validates their identities. In this way, the representation of heritage on Wikipedia recapitulates a long-standing tradition of colonial subordination of one group's heritage and identity by a more powerful and dominant group.

@@

The agency of people participating in heritage-related activities on Wikipedia is highly constrained by the technical characteristics of editing, bots, and consensus-valuing culture. Although there are some niches where editors can debate, concerns about state-like actors, violence & destruction, deal-making, etc. are poorly represented. Typical conflicts about Wikipedia CS-WHL articles are hyper-local, and process-centered. Spatial patterns in the edits to Wikipedia articles about CS-WHL show a kind of digital colonialism of world heritage information, with most edits emanating from the United States. Because of these technical and cultural characteristics of Wikipedia, its articles present a discourse that actively obscures the power relations that give rise to them, and makes opaque debates about how heritage is involved in the production of identity and power [cf.@harvey2001heritage].

That said, Wikipedia is valuable and important as a Big Data source of social information on heritage for several reasons. Although it lacks the velocity of other social media, such as Twitter, it preserves a distinctively complex record of human interactions with the past at simultaneously high scales, high resolution, and in highly structured ways, afforded by no other platform. As a non-profit entity, Wikipedia's data are freely available to researchers, unlike Twitter and Facebook. This has important ethical implications because it eliminates the need for academic collaboration with commercial platforms, whose business interests may not align with the researchers' ethics [@schroeder2014big]. As a context of heritage production, a resource of information on heritage, and an artefact of contemporary digital heritage, Wikipedia is remarkable and holds substantial research potential. Given the tremendous popularity of Wikipedia, is it important that future work on digital heritage in this context continue to critically examine what is absent from, and uncontested in, Wikipedia, and who benefits and suffers from these absences.

>>>>

An additional consideration for understanding participation in online peer production communities are the technical schemas of the software that Wikipedia runs on, MediaWiki. This is a complex toolkit that enables participation in Wikipedia in highly structured ways. On one hand, these structured behaviours produce structured datasets that are well suited to data science methods for efficient computational analysis of large numbers Wikipedia articles. On the other hand, they constrain and limit the agency of the user, canalizing their behaviour into a small number of possible actions and acceptable modes of discourse and engagement with other users [@iba2010analyzing]. While Wikipedia has elements that are ubiquitous on the Internet, such as links on articles that take the user to other articles or other pages on the Internet, it also has several less common elements that contribute to its unique technicity, resulting in very specific types of relationships between human users and the technical elements of the Wikipedia project [@niederer2010wisdom; @weltevrede2016platform]. For example, every edit to an article is tracked in a publicly accessible version control system attached to that article. This exposes the article creation process in highly granular detail; for any given article, we can see how many editors contributed, the size of their edits and their distribution over time, among other things [@priedhorsky2007creating]. Wikipedia has a special category of edit called the 'revert' which allows a user to restore an article to an earlier state to remove recent vandalism (such as the addition of irrelevant or offensive material). This special revert action, combined with a 'talk' page attached to each encyclopedia article for threaded discussion among editors, allows us to detect and study the dynamics of conflicts arising from the creation and editing of articles [@suh2007us]. While articles themselves must be written to conform to the fundamental Wikipedia policy of NPOV (Neutral Point Of View), the talk page is where different views are expressed and negotiated among editors in detail. Taken together, a high proportion of 'revert' type edits, and a high word count on the talk page can indicate the controversiality of an article [@yasseri2012dynamics].

@@

Of the `r nrow(wh_unesco)` cultural sites on the World Heritage List at the time of writing, we found Wikipedia articles for `r nrow(some_page_variables)`. To characterize the overall variability of article according to the basic characteristics described above, we applied a Uniform Manifold Approximation and Projection (UMAP) to reduce the dimensionality of our data [@mcinnes2018umap]. After a two-dimensional projection was obtained, we used Hierarchical Density-Based Spatial Clustering of Applications with Noise (HDBSCAN) to identify clusters [@campello2013density], and Random Forests to identify variable loadings [@breiman2001random] so we could determine which of the basic characteristics were most important in driving variability between the articles. This revealed `r n_clusters` clusters, with the most important variables for determining cluster membership being the number of items in the reference list at the end of the article, the number of words in the article, and the number of Wikilinks out from the target article to other Wikipedia articles. That said, `r n_no_cluster` (`r round(prop_no_cluster * 100)`%) articles were not assigned to a cluster by the HDBSCAN algorithm, indicating that the basic characteristics of CS-WHL articles do not partition the articles in highly structured and discrete groups.

@@

The scatterplot on the lower right of Figure \@ref(fig:wh-wiki-content-and-consumption) shows the values of inward Wikilinks, article views and article word count for all CS-WHL articles. The labeled points in the upper right quadrant are the articles that receive most of the attention. Of the CS-WHL notable for conflict discussed above, only Timbuktu is visible among these highly popular articles. It is also the only site on the List of World Heritage in Danger that is among these highly popular articles. We reviewed the talk page for the Timbuktu article to determine if the attention received by the article might relate to the issues leading to armed conflict at the site. Of the `r timbuktu_talk_page_word_count` words on the talk page, only 168 are on the topic of conflict and destruction, with the editors discussing how to describe the scale of the damage to the temples. The majority of talk page content is about filling in missing detail, suggestions for or notifications of minor corrections. This is also the case for the talk page for Auschwitz, another very popular CS-WHL article. Although Wikipedia articles on this site in other languages have different points of emphasis [@wolniewicz2016framing], the discussion among editors is generally technical and precise, and provocative comments usually ignored rather than directly disputed. This can be contrasted with the talk page for the Holocaust, where @pfanzelter2015crossroads found abundant evidence of conflict among editors.

@@

For the special 'revert' edit type, we see that the proportion of all edits per  CS-WHL articles is similar to other articles, but has a left-skewed distribution indicating a higher number of articles that have very few revert edits (Figure \@ref(fig:wh-wiki-vandal-revert)). We also identified edits with the string 'vandal' in the edit summary as a similar type of edit to the revert edit, e.g. "Edits by 72.49.241.71 identified as vandalism". CS-WHL articles generally have fewer edits about vandalism than our random sample. The shape of the distribution of edits about vandalism has a smaller second mode to the left of the peak, indicating that a large number of CS-WHL articles have very few edits about vandalism (Figure \@ref(fig:wh-wiki-vandal-revert)). Among the CS-WHL articles that have high proportions of reverts and edits about vandalism are highly iconic sites in the Western canon of culture history, e.g. the Sydney Opera House, the Tower of London, and the Statue of Liberty [cf. @harrison2013forgetting]. In reviewing a sample of several hundred reverted edits for each of these, we found that nearly all of them are undoing the addition of short strings of text (e.g. profanities, spam, nonsense). Much of this vandalism is playful, in the spirit of "'I am', a statement that one is present and alive", as @baker2003red described historical graffiti on the Reichstag in Germany by Russian soldiers in WW2. Once again, of the CS-WHL sites with a history of conflict or on the in-danger list, only Timbuktu appears here as having high proportions of revert and vandalism-reversing edits.

@@

Perhaps the most striking spatial pattern in our results is how much of the edits to CS-WHL articles in any part of the world come from the United States. The anonymous editors of articles about CS-WHL are generally not located in the same country as the site they are editing. This suggests a digital colonialism of world heritage information on Wikipedia, where heritage is interpreted and communicated by people who are not part of the descendant community traditionally associated with the site. This is an important observation because it is these editors that control the production of knowledge about CS-WHL articles, and the ability to control information about heritage has several implications. First is that the choices of these editors define community identity, that is, the community of Wikipedia editors, and also the broader community of Internet users that read Wikipedia, by indicating which sites are important and visible, and which are not by contributing to a digital 'authorized heritage discourse' [@smith2012constrained] that prioritizes their own self-interests. Second is that traditional communities associated with CS-WHL have little control over how their own heritage is represented on Wikipedia. We see this spatial pattern on Wikipedia as part of a trend of globalization of heritage, where places that had traditionally contributed to local or regional identities are turned into sanitized playgrounds for rich tourists [@bernbeck2004political]. Wikipedia may be seen as virtual playground for Americans to construct a universal heritage that reflects their neo-liberal cultural values, and validates their identities, perhaps reflecting a cultural rhetorcic 'heritage populism' [@reynie2016specter]. In this way, the representation of heritage on Wikipedia recapitulates a long-standing tradition of colonial subordination of one group's heritage and identity by a more powerful and dominant group.

@@

The agency of people participating in heritage-related activities on Wikipedia is highly constrained by the technical characteristics of editing, bots, and consensus-valuing culture. Although there are some niches where editors can debate, concerns about state-like actors, violence & destruction, deal-making, etc. are poorly represented. Typical conflicts about Wikipedia CS-WHL articles are hyper-local, and process-centered. Spatial patterns in the edits to Wikipedia articles about CS-WHL show a kind of digital colonialism of world heritage information, with most edits emanating from the United States. Because of these technical and cultural characteristics of Wikipedia, its articles present a discourse that actively obscures the power relations that give rise to them, and makes opaque debates about how heritage is involved in the production of identity and power [cf. @harvey2001heritage].

That said, Wikipedia is valuable and important as a Big Data source of social information on heritage for several reasons. Although it lacks the velocity of other social media, such as Twitter, it preserves a distinctively complex record of human interactions with the past at simultaneously high scales, high resolution, and in highly structured ways, afforded by no other platform. As a non-profit entity, Wikipedia's data are freely available to researchers, unlike Twitter and Facebook. This has important ethical implications because it eliminates the need for academic collaboration with commercial platforms, whose business interests may not align with the researchers' ethics [@schroeder2014big]. ALthough free of these commercial interests, a different set of ethics is implied by the exquisite administrative and technical complexity of Wikipedia. These structures serve to normalize and historize inequalities, and manage threats to models of control and expertise about heritage [@harrison2015beyond].  As a context of heritage production, a resource of information on heritage, and an artefact of contemporary digital heritage, Wikipedia is remarkable and holds substantial research potential. While the Neutral Point of View is a central policy for Wikipedia articles, representations and definitions of cultural heritage are never neutral in any discourse, because it always comes from some perspective [@lahdesmaki2019conflicts]. Given the tremendous popularity and visibility of Wikipedia, is it important that future work on digital heritage in this context continue to critically examine what is absent from, and uncontested in, Wikipedia, and who benefits and suffers from these absences.
